---
title: "Assignment 1 - Language Development in ASD - Power and simulations"
author: "[YOUR NAME]"
date: "[DATE]"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome to the third exciting part of the Language Development in ASD exercise

In this part of the assignment, we try to figure out how a new study should be planned (i.e. how many participants?) in order to have enough power to replicate the findings (ensuring our sample size is adequate, our alpha at 0.05 and our beta at 0.8):
1- if we trust the estimates of the current study. Report the power analysis and comment on what you can (or cannot) use its estimates for.
2- if we are skeptical of the current study. Report the power analysis and comment on what you can (or cannot) use its estimates for.
3- if we only have access to 30 participants. Identify the power for each relevant effect and discuss whether it's worth to run the study and why
The list above is also what you should discuss in your code-less report.


## Learning objectives

- Learn how to calculate statistical power
- Critically appraise how to apply frequentist statistical power

### Exercise 1

How much power does your study have (if your model estimates are quite right)?
- Load your dataset (both training and testing), fit your favorite model, assess power for your effects of interest (probably your interactions).
- Report the power analysis and comment on what you can (or cannot) use its estimates for.
- Test how many participants you would have to have to replicate the findings (assuming the findings are correct)

N.B. Remember that main effects are tricky once you have interactions in the model (same for 2-way interactions w 3-way interactions in the model). If you want to test the power of main effects, run a model excluding the interactions.
N.B. Check this paper: https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12504
You will be using:
- powerSim() to calculate power
- powerCurve() to estimate the needed number of participants
- extend() to simulate more participants

```{r}
setwd("C:/Users/louis/OneDrive - Aarhus universitet/AU Onedrive - RIGTIG/- 3. Semester/Experimental Methods III/Classes/Assignment 2/Assignment-2")

# Installing and loading packages
# install.packages("simr", dependencies = T)
# library(pacman)
# p_load(tidyverse, lme4)
# library(simr)

library(githubinstall)
githubinstall("simr", lib = .libPaths())
print(pacman::p_path() == .libPaths())
pacman::p_load(tidyverse, lmerTest, simr)


# Loading data
df_train <- read_csv("train.csv")
df_test <- read_csv("test.csv")

# Making the test set an extension of the training set regarding IDs
df_test$Child.ID <- 66 + df_test$Child.ID

# Merging training and test data
df <- merge(df_train, df_test, all = T)

# Removing NAs
df <- df[-which(is.na(df$CHI_MLU)),]

# Making sure, Child IDs are sequential from 1 and upwards
df$Child.ID <- as.numeric(as.factor(as.character(df$Child.ID)))


# Fitting models
model0 <- lmer(CHI_MLU ~ Visit + Diagnosis + (1 + Visit|Child.ID), df, REML = F, control = lmerControl(optimizer = "nloptwrap", calc.derivs = F))

model <- lmer(CHI_MLU ~ Visit * Diagnosis + (1 + Visit|Child.ID), df, REML = F, control = lmerControl(optimizer = "nloptwrap", calc.derivs = F))

# Checking the main effect estimates
fixef(model0)["Visit"] # main effect of 0.23
fixef(model0)["DiagnosisTD"] # main effect of 0.23

# Checking the interaction effect estimate
fixef(model)["Visit:DiagnosisTD"] # interaction effect of 0.25

# Running the power analysis
set.seed(1) # Setting a fixed "random"

# On the main effects (model without interactions)
powvis <- powerSim(model0, test = fixed("Visit"), nsim = 100, seed = 1)
powvis
# 100% power for detecting effect of visit of 0.23 (nsim = 100)

powdiag <- powerSim(model0, test = fixed("Diagnosis"), nsim = 100, seed = 1)
powdiag
# 30% power for detecting effect of diagnosis of 0.23 (nsim = 100)

# On the interaction effect
powinter <- powerSim(model, test = fixed("Visit:Diagnosis"), nsim = 100, seed = 1)
powinter
# 100% power for detecting the three-way interaction effect of 0.25 (nsim = 100)


# Report the power analysis and comment on what you can (or cannot) use its estimates for.

# The main effect of visit has a great power (100%) for detecting the effect of 0.23 on child MLU. That is, an increase per visit (per 4 months) of 0.23 in child MLU.
# Contrary, the main effect of diagnosis (0.23 on child MLU) is more difficult to detect, because the current study only has 30% power. Meaning, that it is more difficult to detect the average increase of 0.23 morphemes per utterance when going from children with ASD to typically developing children. This would require more participants.
# When calculating power for the interaction effect between "Visit" and "Diagnosis", it results in 100% power. This means, that an interaction effect of this size (0.25) is easily detected with a sample like this. Thus, every time visit increases with 1, the difference between child MLU for ASD children and TD children increases with 0.25.


# Test how many participants you would have to have to replicate the findings (assuming the findings are correct)

# We assume that we do not need to extend our simulation, when it comes to the main effect of "Visit" and the interaction effect between "Visit" and "Diagnosis", since the powers of these effects are 100% with our current number of participants. However, for the main effect of "Diagnosis" an extension (to simulate more participants) is needed in order to detect this effect.

model_extended <- extend(model0, along =  "Child.ID", n = 220)

# Main effects
pc_Visit <- powerCurve(model0, test = fixed("Visit"), along = "Child.ID", nsim = 100, seed = 1)
print(pc_Visit)
plot(pc_Visit)
# In order to replicate the findings of a main effect of Visit, one would need around 9 participants (resulting in power of .80)

pc_Diagnosis <- powerCurve(model_extended, test = fixed("Diagnosis"), along = "Child.ID", nsim = 100, seed = 1)
print(pc_Diagnosis)
plot(pc_Diagnosis)
# In order to replicate the findings of a main effect of Diagnosis, one would need around 196 participants (resulting in power of .80)

pc_inter <- powerCurve(model, test = fixed("Visit:Diagnosis"), along = "Child.ID", nsim = 100, seed = 1)
print(pc_inter)
plot(pc_inter)
# In order to replicate the findings of a three-way interaction effect, one would need around 16 participants (resulting in power of .80)


```


### Exercise 2

How would you perform a more conservative power analysis?
- Identify and justify a minimum effect size for each of your relevant effects
- take the model from exercise 1 and replace the effects with the minimum effect size that you'd accept.
- assess the power curve by Child.ID, identifying an ideal number of participants to estimate each effect
- if your power estimates do not reach an acceptable threshold simulate additional participants and repeat the previous analysis
- Report the power analysis and comment on what you can (or cannot) use its estimates for.

```{r}
# - Identify and justify a minimum effect size for each of your relevant effects

# For main effect of visit
# A minimum effect size for the main effect of visit, could be justified to be 0.16. This is based on the results of a study done on child MLU development (Ranti & Tadris Bahasa Inggris, 2015). According to this article, the increase in MLU from 15-28 months is 0.5. This is a 13-month gap, corresponding to approximately 3 visits in our data. This suggests that the MLU development pr. visit is around 0.16.

# For main effect of Diagnosis
# For the main effect of ‘Diagnosis’, a minimum effect size of 0.07 was specified, based on a study comparing children with ASD and TD children (Swensen, Kelley, Fein, & Naigles, 2007). This study showed an average difference of 0.07 at 21 months of age. This effect size seems absurd and incredibly small, however it was used

# For two-way interaction effect between Visit and Diagnosis
# # # # # # # # # # # # # WRITE HERE # # # # # # # # # # # # # #


# - Take the model from exercise 1 and replace the effects with the minimum effect size that you'd accept.
# Specifying a minimum effect size for the main effects
fixef(model0)["Visit"] <- 0.16
fixef(model0)["DiagnosisTD"] <- 0.07

# Specifying a minimum effect size for the two-way interaction effect
fixef(model)["Visit:DiagnosisTD"] <- 0.1 

# - Assess the power curve by Child.ID, identifying an ideal number of participants to estimate each effect
pc_Visit2 <- powerCurve(model0, test = fixed("Visit"), along = "Child.ID", nsim = 100, seed = 1)
print(pc_Visit2)
plot(pc_Visit2)
# In order to replicate the findings of a main effect of "Visit" with a more conservative approach (effect size of 0.16), one would need around 14 participants (resulting in power of .80)

# Extending the model even more in order to be able to detect this very small effect
model_extended2 <- extend(model0, along =  "Child.ID", n = 2200)

# Checking the effect size is as we have specified in the newly extended model
fixef(model_extended2)

pc_Diagnosis2 <- powerCurve(model_extended2, test = fixed("Diagnosis"), along = "Child.ID", nsim = 100, seed = 1)
print(pc_Diagnosis2)
plot(pc_Diagnosis2)
# In order to replicate the findings of a main effect of "Diagnosis" with a very conservative approach (effect size of 0.07), one would need around 2200 participants (resulting in power of .80)

# A smaller extension is needed to detect this effect
model_extended3 <- extend(model, along = "Child.ID", n = 80)

# Checking the effect size is as we have specified in the newly extended model
fixef(model_extended3)

pc_Inter2 <- powerCurve(model_extended3, test = fixed("Visit:Diagnosis"), along = "Child.ID", nsim = 100, seed = 1)
print(pc_Inter2)
plot(pc_Inter2)
# In order to replicate the findings of an interaction effect with a more conservative approach (effect size of 0.1), one would need around 59 participants (resulting in power of .80)
```


### Exercise 3

Assume you have only the resources to collect 30 kids (15 with ASD and 15 TDs). Identify the power for each relevant effect and discuss whether it's worth to run the study and why

```{r}
# Main effects
pc_30_Visit <- powerCurve(model0, test = fixed("Visit"), along = "Child.ID", nsim = 100, breaks = seq(from = 1, to = 30, by = 5), seed = 1)
print(pc_30_Visit)
plot(pc_30_Visit)
# With 30 children, the power for the main effect of "Visit" is 100%.

pc_30_Diagnosis <- powerCurve(model0, test = fixed("Diagnosis"), along = "Child.ID", nsim = 100, breaks = seq(from = 1, to = 32, by = 2), seed = 1)
print(pc_30_Diagnosis)
plot(pc_30_Diagnosis)
# With 30 participants, the power for the main effect of "Diagnosis" is around 5%.

# Interaction effect
pc_30_inter <- powerCurve(model, test = fixed("Visit:Diagnosis"), along = "Child.ID", nsim = 100, breaks = seq(from = 1, to = 32, by = 2), seed = 1)
print(pc_30_inter)
plot(pc_30_inter)
# With 30 participants, the power for the interaction effect is 44%.



# # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # #



# Creating a random sample of the data to test the power with 30 participants (15 ASD and 15 TD).
set.seed(1)
DATA = df %>% group_by(Child.ID, Diagnosis) %>%
  summarise(count = n()) %>%
  filter(count == 6)

ASD <- DATA[DATA$Diagnosis == "ASD",]
ASD_unique = sample(as.vector(unique(ASD$Child.ID)), 15)
ASD <- df[df$Child.ID %in% ASD_unique,]

TD <- DATA[DATA$Diagnosis == "TD",]
TD_unique = sample(as.vector(unique(TD$Child.ID)), 15)
TD <- df[df$Child.ID %in% TD_unique,]

df30 <- bind_rows(TD, ASD)
summary(as.factor(df30$Diagnosis))


# Specifying models
model0_30 <- lmer(CHI_MLU ~ Visit + Diagnosis + (1 + Visit|Child.ID), df30, REML = F, control = lmerControl(optimizer = "nloptwrap", calc.derivs = F))

model_30 <- lmer(CHI_MLU ~ Visit * Diagnosis + (1 + Visit|Child.ID), df30, REML = F, control = lmerControl(optimizer = "nloptwrap", calc.derivs = F))

# Specifying a minimum effect size for the main effects
fixef(model0_30)["Visit"] <- 0.16
fixef(model0_30)["DiagnosisTD"] <- 0.07

# Specifying a minimum effect size for the two-way interaction effect
fixef(model_30)["Visit:DiagnosisTD"] <- 0.1 

# Calculating power
# On the main effects
powvis_30 <- powerSim(model0_30, test = fixed("Visit"), nsim = 100, seed = 1)
powvis_30
# 98% power for detecting the specified effect of visit (nsim = 100)

powdiag_30 <- powerSim(model0_30, test = fixed("Diagnosis"), nsim = 100, seed = 1)
powdiag_30
# 11% power for detecting the specified effect of diagnosis (nsim = 100)

# On the interaction effect
powinter_30 <- powerSim(model_30, test = fixed("Visit:Diagnosis"), nsim=100, seed = 1)
powinter_30
# 25% power for detecting the specified interaction effect (nsim = 100)

```
